---
title: "Central tendency measures"
output: html_notebook
---

note that in the R codes `x` is a vector of values. Vectors can be created with `c()` for examle: <br>
```{r}
c(1,2,3,4,5,6)
```

### Mode
- A simple central tendency that can be used for all scale levels
- The mode indicates which expression occurs most frequently
- If the absolute (or relative) frequency distribution has a unique maximum, the mode is unique and is
denoted by x(mod)
- In the bar chart, the mode is the expression with the highest column
- the mode does not always have to be unique
- calculation in R:
`Mode <- function(x) { ux <- unique(x) ux[which.max(tabulate(match(x, ux)))] } mode <- Mode(x)`

##### Properties of the mode <br>
- If only very few values of the list of observations match the mode is not very meaningful

---

### Arithmetic mean
- calculation in R: `mean(x)`
- for grouped data the arithmetic mean from the midpoints of the classes has to be calculated <br>

##### Properties of the arithmetic mean <br>
- the arithmetic mean is very sensitive to outliers
- This sensitivity to outliers can be undesirable (e.g. individual errors in the data lead to large changes in the arithmetic mean)

---

### Median
- for at least ordinally scaled characteristics
- One half of the values of the original list should be below the median and the other half above
- For odd 𝒏𝒏 the median is the middle value of the ordered data
- calculation in R: `median(x)`

#### Properties of the median
- robust against outliers
- At least 50% of the values in the data are less or equal to the median
- At least 50% of the values in the data are greater or equal to the median

---

### Quantiles
- represent central tendency measures that also already provide information on the dispersion of the data
- Let 0 < p < 1. For any value xp for which
    - Let 0 < p < 1. For any value xp for which
    - at least a proportion 1 − p is greater or equal than xp
- calculation in R: Quantiles 0.2, 0.25, 0.5, 0.75 <br>
`c(xordered[floor(n * 0.20) + 1], xordered[floor(n * 0.25) + 1], xordered[floor(n * 0.5) + 1], xordered[floor(n * 0.75) + 1])`
- calculation of interquartile range <br>`(xordered[floor(n * 0.75) + 1]) - (xordered[floor(n * 0.25) + 1])`

#### Properties of the Quantiles and the interquartile range
- robust against outliers
- do not contain any information about the left and right ends

---

### five point summary
- The five-point summary of the distribution of ametric variable consists of : <br> x(min), 0.25 - Quantile, median, 0.75 - Quantile, x(max)
- visualization in R: `boxplot(x)`
- example:
```{r}
x <- c(1,4,6,3,7,9,4,2,1,4)
boxplot(x)
```

---

#### Overview

|                                       |          | Mode | Arithmetic mean | Median | Quantiles |
|---------------------------------------|----------|------|-----------------|--------|-----------|
| Applicable for different scale levels | nominal  | yes  | no              | no     | no        |
|                                       | ordinal  | yes  | no              | yes    | yes       |
|                                       | cardinal | yes  | yes             | yes    | yes       |
| Robust against outliers               |          | yes  | no              | yes    | yes       |

[Back to Table of Contents](../)